\documentclass[letterpaper,titlepage,twoside]{article}
\usepackage{xparse}
\usepackage[left=2in,right=2in,top=2in,bottom=2in]{geometry}
\usepackage{hyperref}
%\usepackage{navigator}

\NewDocumentCommand{\word}{v}{\{~\texttt{#1}~\}}
\begin{document}
%\outline[open]{3>}[outlinelocation]{VTIL}
\begin{titlepage}
  \begin{center}
    \vspace*{2in}
    \Huge
    VTIL\\
    \LARGE
    A VAX Threaded Interpretive Language
    \vspace*{1in}
    Thomas Kurt Bond

    \vspace*{0.5in}
    Prof. M. Berry, Chairman\\
    Math and Computer Science Department

    \vspace*{0.5in}
    4 March 1988
  \end{center}
\end{titlepage}

\begin{abstract}
  Low level languages such as assembly are difficult to work with and
  machine specific.  High level languages are much more restrictive,
  allowing little or no expansion of the language, and often limit the
  programmer's access to the real machine.  Compiled languages are
  efficient, but corrections and changes require much time and
  interactive debugging is limited at best.  Interpreted languages are
  easy to use and debug, but very inefficient.  One useful compromise
  between these opposing categories is the class of threaded
  interpretive languages{\bf TIL}s.  They have many advantages.  {\bf
    TIL}s incude such elements of structured programming as
  one-entry/one-exit modules; structured control statements such as
  if---then---else, looping statements; and no goto statement, but also
  allow the programmer to acces the real machine and to extend the
  language in a portable manner.  They are interactive, allowing easy
  debugging and program modification, yet are more efficient in
  execution time than interpreted languages and in some cases nearly as
  efficient as compilers, and are also very efficient in code space.
  They provide a flexible but portable method for a programmer to build
  a customized programming environment and are also much easier to
  implement than a compiler or interpreter.  There are many different
  examples of {\bf TIL}s, but the best known are Forth and Stoic.  The
  primary characteristics of a {\bf TIL} are as follows.  The {\it
    dictionary} is where the definitions of definitions of all {\it
    words} (which are similar to procedures and functions) that have
  been defined are stored.  All words take their arguments from and
  return values on, the {\it stack}.  A {\it return stack} contains
  information about the execution of the current word and loop.  Global
  constants and variables can be defined and local variables are
  available (by use of the stack inside words).  The author proposes to
  develop a simpl {\it TIL} with all the capabilities needed to allow
  programs to be developed.  This would include arithmetic, logical, and
  relational operators; words to allow the defining of constants,
  variables, and new words; input and output to the terminal and disk
  blocks; control structures including if---then---else and looping
  constructs; and a simple line mode editor that works on disk blocks.

  {\bf Key Words:}  Threaded Interpreted Languages, Forth Stoic.
\end{abstract}

\tableofcontents
\newpage

\begin{center}
  \Large
  VTIL\\
  A VAX Thread Interpretive Language
\end{center}


\section{Introduction}

VTIL is a threaded interpretive language impmlemented on the VAX.  It
is an interactive interpreter/compiler for a stack based langauge.
VTIL includes arithmetic, logical, and relational operators, though
only integer arithmetic is provided.  String Manipulation procedures
are included, as are input and output to the terminal and to disk
block files (which reside in a block-file, since VTIL runs on top of
VMS).  Source code and commands can be entered from the terminal or
from text files.  As an example of a VTIL application, I have written
a small line-oriented text editor.  VTIL is a general purpose
programming language, and is suitable at its current level for most
programming tasks not including extensive file access.

My main purpose in writing VTIL was to explore the implementation of a
threaded interpretive language.  I am very interested in the design
and implementation of languages, but the task of writing a compiler or
interpreter for a reasonably power language seemed impractical given
my limited experience and the amount of time allotted for the seminar.
However, the class of languages known as Threaded Interpretive
Langauges (hereafter called TILS) appeared not only to be much easier
to implement than conventional languages but also to have many
interesting features, such as combining the interactive nature of an
interpreter with execution speed almost as greatas a compiler in many
cases.  I had originally intended to write a {\em pointer}-threaded
TIL, but the advantages of a {\em subroutine}-threaded TIL looked
extremely intriguing, especially since I was writing in VAX-Macro,
which is especially suited to subroutine threading.  A secondary
purpose in this seminar was investigating the interesting
characteristics of the TIL class of languages, such as their use of
a data stack as the primary method of communication between routines,
the extreme extensiblity of the langauge, and the dictionary of words.

Becuse many of the names of routines in VTIL contain punctuation, to
avoid confusion in this paper all the VTIL names, or {em words}, will
appear in a different type face and inside braces.  An example is the
word \word{b!}, which will store a byte value in a variable.  Braces
will never appear in any VTIL words in this manual.  Also, within a
sequence of VTIL words, the string `...' indicates a series of VTIL
words that were omitted as irrelevant.  Again, none of the VTIL words
discussed in this paper will contain `...' as part of their name.

\section{Concepts}

VTIL and TILs in general use many concepts that are alien to
programmers who deal mainly with languages like FORTRAN or Pascal.
TIL programming benefits from a whole new style, and many programming
habits from more conventional languages are ill-suited to TILs.  While
some adjustment to this new programming style is needed, it is amply
rewarded.  Because the fundamental concepts are used throughout the
language I will first explain the most important ones.

The syntax of VTIL is extremely simple: the input stream is a sequence
of tokens separated from each other by blank spaces (tabs are {em not}
identical to blanks).  Each token is either a {\em word} or data for a
word.  The advantage of of this is that no complicated parsing of the
input needs to be done.  If a more complex syntax is needed, the
programmer can easily add it to his application, but usually it is not
needed.

Unlike most languages, VTIL is stack-based and uses postfix notation
for almost everything, instead of the more familiar infix (or
algebraic) notation.  That is, the data for an operation always
precedes the operator.  The advantage of this is that all routines
take their input data from the same place the {em data stack},
eliminating any need for a complicated calling sequence for routines,
such as \verb|call doit (x,y*x,z)|, which is common to both Fortran
and Basic, with all of the overhead involved in setting up such a
call.  Because the data stack is used so often, it will usually be
referred to simply as the stack.

In addition to the data stack VTIL also makes use of two other
stacks.  The first is the {em loop stack}, which is used primarily
during execution by the various looping words such as \word{do ... loop} 
which maintain loop indexes and limits on the stack, and
during compilation by words that compile control structures, such as
\word{if ... else ... endif}, which need to store information from the
word that starts the control structure until the execution of the word
that ends it.  A loop with an index maintains its index on the stack,
and the top loop stack item inside such a loop is the index, which can
be copied to the data stack to be used for computation.  The loop
stack is also useful for temporary storage of data that would
otherwise be in the way on the data stack.  When the loop stack is
used for temporary storage, however, one has the remove the items on
it at the same level of nesting as they were placed on it; that is,
inside the same lop.  Otherwise the indices of that loop and any
containing loops will be corrupted.  The final stack that VTIL makes
use of is the {em return stack}, which is actually the normal VAX
system stack.  It contains the address to resume execution at when the
current word is finished executing, just as normal.   Since the system
stack is used normally it is easy to set up VTIL to call any VAX
run-time library routines.

In VTIL the fundamental unit is the {\em word}.  Unlike other
languages, where constants are distinct from variables which are
distinct from functions and procedures, {em everything} in VTIL is a
word.  A constant in VTIL is a word which pushes a value to the
stack.  A variable in VTIL is a word which pushes the {\em address} of
a value to the stack.  The closest VTIL gets to the procedure or
function of other languages is the {em colon definition}, which
executes the words that make it up when it is executed.  Words can be
written that declare data types or declare and initialize variables.
All words are invoked in exactly the same way.

One of the central concepts of a TIL is the {em dictionary}.  The
dictionary is a group of linked lists containing al the defined words
whether they are defined by the user or were provided with the
language.  Regardless of how they were defined, they are all treated
exactly the same.  When a token is to be executed or compiled, the
dictionary is searched for a word with a matching name.  If the word
is found, it is executed or compiled; otherwise an error message is
written to the terminal for the user.  It is possible to define a new
word with the same name as an old word, but the new version of the
word will only be used by words defined after it is; words defined
before it will continue to use the old version, since they have
already been compile to call that version of the word.

An important concept in a TIL is that of the {\em vocabulary}.  A
vocabulary is one of several independent linked lists that are woven
through the dictionary, only one of which is searched for words at a
time.  Vocabularies are used to separate words into groups according
to function.  For instance, all words that take make up the editor
vocabulary can be completely separated from the words that make up the
vocabulary for a text formatter.  One can therefore have words with
the same names, but a different meaning, in different vocabularies,
and only have the word in the proper vocabulary found when working in
that vocabulary.  Usually vocabularies are linked at their ends into
the {\em kernel} vocabulary, which contains the words that were
provided by the language.

The distinction between the dictionary and vocabularies is important:
the dictionary can be considered as a contiguous area of memory in
which all words reside, while a vocabulary is a linked list of words
in the dictionary which may be intertwined with other vocabularies,
but still independent of them.  When one speaks of searching the
dictionary, actually one means search one of the vocabularies in the
dictionary.  A good analogy for the dictionary and the vocabularies is
a rope: each individual strand or thread in the rope is like a
vocabulary, but the whole intertwined bundle that makes up the rope is
a dictionary.

\section{Characteristics and Capabilities}

VTIL is a general purpose language that can be used for many things.
It is easy to program and to debug programs in.  This section
discusses the major characteristics and capabilities of VTIL.

\subsection{Interactive use}

To use VTIL you need to run the VTIL interpreter/compiler,
\verb|VTIL.EXE|.  When you have done so, VTIL will print its prompt,
\verb|VTIL>|, and wait for you to enter commands.  To exit VTIL you
can type \verb|ctrl/z| or \word{exit} at the \verb|VTIL>| prompt.  To
execute a VTIL word (like \word{exit} to exit the language) you just
type the letters of that make up the name of the word, preceded and
followed by any number of blank spaces (tabs are not allowed outside
string constants).  Since VTIL is case sensative, you must type its
commands in the same case as they are given in this paper, usually
lower case.  To enter a number (which must be an integer) into the
data stack just type the number, which can have a leading minus sign.
VTIL will not actually execute the word or enter the number into the
stack until you hit \verb|<return>|.  You can place as many words or
numbers on a line as you wish.  Source code and commands can be also
loaded from a text file by giving the command \word{load <filename>},
where \verb|<filename>| is the name of the file to be loaded.  The EVE
command editor can be invoked from inside VTIL as well, by giving the
command \word{eve <filename>}.

If you enter an unknown word, the offending word is echoed to the
terminal and the line it was in is also echoed, like this:
\begin{verbatim}
Unknown token { xxxx_I'm_unknown } in line:
1 2 3 xxxx_I'm_unknown xyz 1 3 + .
\end{verbatim}

If source code was being loaded from a file, the line number and
filename are also displayed.  After an error the rest of the input
stream is discarded and the data and loop stacks are cleared.

If you have started a command that is taking a long time, you can
interrupt it by typing \verb|ctrl/c|, which will return you to the
\verb|VTIL>| prompt (this is useful for stopping infinite loops).
VTIL will aso catch access violations and most other errors, but there
is a builtin limit to the number of errors it will catch.  After a
certain number of errors and \verb|ctrl/c|'s are trapped, VTIL will
abort with a message indicating that the maximum number of errors has
been exceeded.  This is intended to stop an unattended program with
errors instead of allowing it to continue indefinitely.

\subsection{Simple Arithmetic and the Stack}

The word \word{.} can be used to see the value of the top of the
stack.  However, \word{.} pops the top of the stack, so the number is
lost after it is used; most VTIL words consume their arguments.  The
word \word{.sb} prints the contents of the stack from the bottom up
without destroying its contents, while the word {.st} prints the
contents of the stack from the top down without destroying its
contents.  The words for addition, subtraction, multiplication,
division, and modulus are supplied and the expected mathematical
symbol as their name (except for modulus, which is \word{mod}).
Another arithmetic word is \word{/mod} which divides the second stack
item by the top stack item and leaves the quotient and remainder on
the stack, the remainder on top.

\subsection{Comments}

There are two type of comments in VTIL.  The first is called ``skip
line'' and starts with a \word{\} and continues until the end of the
line.  The \word{\} must be separated from other text by at least
one space on each side.  The second type of comment, called an
``inline comment'' or a ``stack comment'', starts with a \word{(}
and ends with the end of the line or a \word{)}, whichever comes
first.  The initial \word{(} must be separated from other text by at
least one space on each side.  This type of comment is usually used
for comments that are embedded between words.

\subsection{Data Types}

VTIL provides very few data types to begin with, but new ones are
extremely easy to define.  The data types provided by VTIL are {\em
  long} (a VAX longword, four bytes), {\em short} (a VAX memory word,
two bytes), {\em byte}, {\em byte-counted string} (a byte containing
the number of characters in the string followed by bytes containing
the characters), and {\em word-counted-string} (a VAX memory word
containing the number of characters in the string followed by bytes
containing the characters).  Words for accessing and manipulating each
of these data types is included in the VTIL kernel, as are words for
defining arrays and other data types.

\subsection{Stack Manipulation Words}

Because VTIL uses the stack to store so much information, it has many
words for rearranging the contents of the stack.  For instance, since
most words consume their arguments, a value must often be duplicated
so the original value can be used later while a copy is used now.
This can be accomplished with the word \word{dup}.  Following are the
most important stack manipulation words.
\begin{description}
  \item[+rot ( n1 n2 n3 -- n3 n1 n2 )]  Rotate the top three items of
    the stack up, bringing the second item to the top and moving the
    previous top to the third entry.
  \item[-rot ( n1 n2 n3 -- n2 n3 n1 )]  Rotate the top three items of
    the stack down, bringing the third entry to the top and pushing
    the other two entries down.
  \item[drop ( n -- )]  Drop the top of the stack.
  \item[dup ( n -- n n)]  Duplicate the top of the stack.
  \item[over ( n1 n2 -- n1 n2 n1 )]  Duplicate the second item of the
    stack on top of the stack.
  \item[swap ( n1 n2 -- n2 n1 )]  Swap the top two items on the stack.
\end{description}

Of course, new stack manipulation words can be defined with these and
other VTIL words if they are needed.  One to consider writing is
\word{under} which is sort of the reverse of \word{over}.  Its stack
comment is \verb|( n1 n2 -- n2 n1 n2 )|.

\subsection{Defining New Words}

Starting with a small set of basic capabilities and building the
capabilities needed, and only those capabilities, from the initial set
is one of the important concepts of VTIL and all threaded interpretive
languages.  Programming in VTIL is done by defining new words and
combining old words into new words.  Because defining new words is so
important in VTIL the words that define words (called {\em defining
  words}) are also extremely important.

All words are invoked in the same way, regardless of what they do: by
their name.  The most common types of words are constants, variables,
and colon definitions.  Names of words can contain any character;
except a blank: however, it is not a good idea to use any nonprinting
character, such as a tab or other control character (indeed, in later
versions of VTIL such names may become illegal altogether).  This
flexibility allows extremely meaning full names, such as
\word{all-here?}, but it also allows extremely unmeaningful names,
such as \word{-&]*!#}.   If names are well chosen the resulting source
code is often very English-like.

\subsection{Constants and Variables}

The form of a constant definition is \word{<n> longconst <name>} which
creates a word \verb|<name>| that pushes <n> on the stack whenever it is
executed.

the form of a variable definition is \word{<n> long <name>}, which
creates a word \verb|<name>| that pushes the address of a long
variable to the stack whenever it is executed.  The variable is
initialized to \verb|<n>|.  The word \word{@} (pronounced fetch) is
executed to obtain the value of a variable.  \word{@} fetches a long
value from memory at the address on top of the stack.  Its stack
comment is \verb|( a -- n)|.  To store a value in a variable the word
\word{!} (pronounced store) is used.  \word{!}'s stack comment is
\verb|( n a -- )|, where \verb|<n>| is the value to store, and
\verb|<a>| is the address to store it at.   An example of the use of
variables and \word{!} and \word{@} follows.
\begin{verbatim}
x @ 1 + x !
\end{verbatim}
This sequence adds one to a variable.  First the address of the
variable is pushed on the stack (by \word{x}) and then the value of
the variable is fetched (by \word{@}).  After that, one is added to
the value (by \word{1 +}), the address of the variable is pushed on
the stack again, and finally the value is stored into the variable (by
\word{!}).

If this procedure seems clumsy, do not worry; variables are not nearly
as important in VTIL as in other languages since all calculation is
done on the the stack, eliminating the need for temporary variables.
Only the initial value and the the final result need variables, and
then only at the beginning or end of the calculation.  Indeed,
variables are often not needed even then, since the calculation will
most likely be encapsulated in a word, which will receive its initial
data from the data stack and leave its results on the stack for
immediate use.  However, VTIL is flexible, so in those cases where a
variable is required, incrementing a counter or some such, there are
other, more convenient ways to deal directly with the value of a
variable.  For instance the sequence \word{3 x +!} adds three to the
value of x, and words like \word{1+!}, which adds one to the value of
any variable, are easy to write if they are needed.

There are also words \word{byte} and \word{short} that define byte and
short integer variables.  The words \word{b@} and \word{b!} are used to
access byte variables, and the words \word{s@} and \word{s!} are used
to access short variables.  The data stack is a long wide, but byte
and short variables are automatically extended to the correct size
when they are placed on the stack by ``zero extending'' them.  That
is, bytes and shorts are considered to be unsigned numbers when they
are placed on the stack, and are always positive.  The words
\word{bs@} and \word{ss@} work just like \word{b@} and \word{s@}
except they extend them to the correct size by ``sign extending'' them
and considering them to be signed numbers instead.

\subsection{Colon Definitions}

Colon definitions are collections of words that execute one after
another, just as procedures and functions in other languages are
collections of statements that execute one after another.  The form of
a colon definition is \word{: <name> ... ;}, where \verb|name| is the
name this word will have.  They start with the word \word{:}, which is
why they are called colon definitions.  The colon is followed by the
name for this new word, which is has the same rules as names for any
other type of word.  Following the name is the list of words that make
up the definitions, terminated by \word{;}.  As an example colon
definition, here is how \word{+!} might be written if it were not
supplied with the langauge.
\begin{verbatim}
: +!  ( n a -- )
   swap over @    \ get the value of the variable
   +              \ add the values
   swap !         \ get the address of variable back on top
                  \ and store value there
; \ end of +!
\end{verbatim}

\subsubsection{Compile-time versus Run-time}

VTIL has two different operating modes: execute mode and compile
mode.  In execute mode, the words in the source are immediately
executed.  In compile mode, the words in the source are compiled into
the dictionary.  VTIL normally is in compile mode only after a
\word{:} and before a \word{;}.  Some words, however, are in a special
vocabulary called the compiler vocabulary which is only searched when
VTIL is in compile mode.  These words are called ``compile-time''
words because instead of being compiled into the dictionary when VTIL
is in compile mode, they are executed.  That is, they are executed
while the word is being compiled, at ``compile-time.''  All other
words are executed at ``run-time,'' when the word is executed.

Compile-time words are needed to provide some way to compile control
structures.  All the loop words, such as \word{do ... loop}, and the
other control structure words, such as \word{if ... else ... endif}
are compile-time words.  These words set up the loops or the control
structures in the word being defined, but are not themselves compiled
into the word.  Usually compile-time words compile some information
into the word being defined that is figured out during compilation
rather than during execution, when that information is has to be
used.  A word is made a compile-time word by following its definition
with the word \word{compiletime}, which moves the most recently
defined word in the \word{current} vocabulary to the compiler
vocabulary.

Note that \word{;} itself has to be a compile-time word, because it
has to execute to switch VTIL from compile mode to execute mode.  If
it were not executed during compile mode, there would be no way to
exit compile mode.  Also, sometimes during a colon definition you want
to execute words that would normally be compiled.  To do so you
enclose them between \word{[} and \word{]}.  \word{[} temporarily
turns off compile mode and \word{]} turns it back on.

\subsubsection{Control Structures}

VTIL includes a full complement of control structures, which may only
be used in colon definitions since they are all compile-time words.
In the following list \verb|<flag>| is a true or false on the top of
the stack, \verb|limit| is the supper or lower limit on a counting
loop, \verb|<index>| is the starting value of the index for a counting
loop, and \verb|<increment>| is the increment to add to the index.
\begin{verbatim}
<flag> if ... else ... endif
<flag> while ... endwhile
<limit> <index> do ... loop
<limit> <index> do ... downloop
<limit> <index> do ... <increment> +loop
repeat ... <flag> until
repeat ... <flag> if ... again
\end{verbatim}

The \word{repeat ... if ... again} loop is especially interesting.
The code begin the \word{repeat} and the \word{if} is always executed
at least once.  If the \verb|<flag>| is true the code between the
\word{if} and the \word{again} is executed and control returns to the
code after the \word{repeat} and before the \word{if}.
If the \verb|<flag>| is false control skips to the
code after the \word{again}.  When the value that controls a
\verb|while <condition> do| loop in Pascal cannot be actually
calculate in the \verb|<condition>| part of the code to calculate it
has to be included before the body of the loop and as the last thing
in the loop, both copies of which must be modified if any changes are
made later.  The \word{repeat ... if ... again} loop allows that code
to be executed before the the loop body before each iteration, so only
one copy of the code needs to be maintained.

\subsubsection{Recursion}

VTIL also allows recursion, as one would expect from a stack based
language.  Because words and can call words defined earlier with the
same name, a special compile-time word \word{recurse} is used in  a
colon definition to call the word being defined.  Following is an
example of the use of recursion, the word for calculating factorials.
\begin{verbatim}
: factorial
   dup 0 <= if        \ for safety, -n! = 1
      drop            \ drop duped 0
      1               \ 0! = 1
   else                      
      dup 1 - recurse \ (n-1)!  call this word recursively
      *               \ n * (n-1)!
   endif
; \ factorial
\end{verbatim}

If \word{factorial} had been used in place of \word{recurse} VTIL
would have looked for an earlier definition of \word{factorial}, not
the one being currently defined, and not finding it would have aborted
this definition with an error.

\subsubsection{Vocabularies}

Vocabularies are subsets of the dictionary.  They allow the programmer
to have different contexts for different purposes, so that \word{x} in
one vocabulary does one thing, but \word{x} in another vocabulary does
something else.  The word \word{vocabulary}, used in the form
\word{vocabulary <name>}, creates a new vocabulary with the name
\verb|<name>|, which is invoked with called \verb|<name>|.  When the
word \verb|<name>| is executed it sets the \word{context} vocabulary
to \verb|<name>| vocabulary, which will now be searched for words to
execute or compile.  Vocabularies created by \word{vocabulary} are
linked back into the ``kernel'' vocabulary (which has all the runtime
words that VTIL starts out with in it) just before the word
\verb|<name>|, so that if the word is not found in vocabulary
\verb|<name>| the kernel vocabulary is also searched.  This allows
using all of the kernel vocabulary words even when in another
vocabulary.  The sequence \word{<name> definitions} changes the
\word{current}
vocabulary (the vocabulary that new words are linked to when they are
defined) so that all the new definitions are compiled into vocabulary
<name> instead of the previous \word{current} vocabulary.
\word{context} can be switched back to the kernel vocabulary by
executing \word{kernel} and \word{current} can be switche back to the
kernel vocabulary by executing \word{kernel definitions} (which also
sets \word{context} to the kernel vocabulary).

\subsubsection{Inline Assembly Code}

VTIL includes its own assembler, and allows you to use assembly
language anywhere in a colon defintion (though ti does not {\em
  encourage} using assembly langauge scattered throughout an
application).  There are many situations where some capability of the
particular machine one is using is needed in a program, but can only
be reached by using assembly langauge.  If this capability is
encapsulated in a word or group of words that act like normal words to
the rest of the program, only these few words will need to be changed
if the program must later be moved to another machine.  If the whole
program were written in assembly language, instead of just a few
sections, the task of transporting it would be much more difficult.
Allowing assembly language anywhere in a colon definition is the way
VTIL provides for exploiting the capabilities of a specific machine to
the fullest, yet still hiding that machine specific code from the rest
of the application.  This also allows innermost loops of programs to
be coded efficiently in assembly language, improving the efficiency of
programs, when such is required.

\subsubsection{Defining Words That Define New Words}
Words like \word{long} and \word{longconst} are defined using two
special compile time words: \word{defer<} and \word{>defer}.  Words
between \word{defer<} and \word{>defer} in a colon definition are not
executed when they are compiled into the word which is at {\em that}
point being defined, and will be executed when {\em that} word is
executed.  Here is an example:
%%% This should have used colon words instead of assembly.
\begin{verbatim}
: 4byte ( n -- ) \ define a word that defines 4byte variables
   create        \ enclose next token in dictionary
   \ compile this into word when *this* word executes
   defer<
      \ move value of variable onto stack
      MOVAB b^r b^ 2 -(ds)
      RSB   \ return from wordd
      \ return to compiling into word to be executed
   >defer
   \ save space for the variable 4 bytes long
   \ and initialize it to <n>
   MOVL (ds)+ (edp)+
;
\end{verbatim}

This example shows how a word with the same effect as \word{long}
might be defined.  Whenever {4byte} is executed a new word is created
with the code between \word{defer<} and \word{>defer} as its code body
and space in the dictionary is allocated for the data and
initialized.

In essence, the compilation of the sequence of words between
\word{defer<} and \word{>defer} is {\em deferred} until the word that
is being defined is executed, at which point those words are compiled
into whatever word is then being defined.  Obviously all words with
\word{defer< ... >defer} in them have to be compile time words.

\subsection{Block Input/Output}

I wanted to allow the VTIL some access to mass storage, yet avoid the
difficulties of dynamic file creation and access.  I therefore decided
that VTIL would only access a single file, called the {\em
  block-file}, and that access would be extremely simple, emulating
the virtual block concept of earlier TILs.  A single file, pointed to
by the logical name VTILBLOCKS (with the default name
\verb|VTILBLOCKS.DAT|), would be created by a separate VAX-Macro
program, and opened, accessed, and closed by VTIL words.  Conceptually,
the block-file is simply one hundred blocks, each 1024 bytes in
length, which can individually read into buffers in VTIL and saved if
any changes are made in them.

To use the block-file in VTIL it must first be opened by
\word{block-open}.  \word{<n> block} reads block \verb|<n>| into a
buffer and leaves the address of that buffer on the stack.  There are
several buffers (the actual number is controlled by a symbol in the
VAX-Macro source code) and each time \word{block} is executed the
block is read into a new buffer, until all the buffers are full.  At
that time a search is made for a buffer that has not been modified,
into which the new block is loaded if found.  If an unmodified buffer
is not found one of the modified buffers is written out to the block
file and that buffer is then used for the new block.  The word
\word{update} marks the most recently accessed block as modified.
After one is done working with the block-file, all modified blocks can
be saved with \word{save-buffers} or abandoned with
\word{empty-buffers}.  The block-file can then be closed with
\word{block-close}.

\subsection{Editor}

As an example of a VTIL application I wrote a simple line mode editor
which operates on blocks from the block file.  It interprets a block
as sixteen lines of 64 characters each, with shorter lines padded with
spaces or nulls.  The editor uses variables, constants, colon
defintions, vocabulary words, and words for string manipulation and
input and output to the terminal.  It coexists with VTIL once it is
has been loaded, so one can issue VTIL commands while editing.  It
includes commands to switch to the editor vocabulary, edit a block,
exit the editor and save any changed blocks, exit the editor without
saving any blocks, clearing lines, and moving lines around.  Also
included are commands to find a text string in the block and replace a
found stext string with another string.  The found strings and the
inserted strings are saved so that the commands can be reissued
without typing anything more than the commands themselves.

\section{Implementation}

My main interest in this seminar was seeing how a TIL is implemented.
I found that a TIL can be implemented quickly: after researching TILs
for a month and thinking about the implementation for two weeks or
sow, I was able to code an incomplete but working version of VTIL in
six days.  This version allowed definition of constants, variables,
and simple colon definitions, including the \word{if ... else ... endif} 
control structure, and did very simple input and output to the
terminal.  The final version of VTIL includes many more capabilities.
There are many interesting things about the implementation of VTIL
what I will discuss next.

\subsection{The VAX and VTIL}

All TILs make heavy use of stacks and VTIL is no exception.
Implementing a TIL on a machine that has a good set of instructions
and addressing modes for implementing stacks makes the TIL writer's
task much, much easier and makes the TIL run much more efficiently.
This is not to say that implementing a TIL on a machine without such
instructions and addressing modes is extremely difficult, and that
such a TIL would be extremely inefficient, for many TILs have been
implemented on machines that do not normally support stacks with ease
and have still proven to be reasonably easy to write and reasonably
efficient.

The VAX instruction set and addressing modes are especially well
suited to implementing a TIL.  Since eleven of the VAX's sixteen
general purpose registers are available to the programmer for use at
all times each stack can have its own dedicated register.  The
addressing modes register deferred, autodecrement, autoincrement,
autoincrement defered, and the various displacement and displacement
deferred addressing modes all make access to the values in a stack
extremely easy and efficient.  For instance, to use th evalue of
whatever in the third item down in the data stack in VAX-Macro, I
could use the operand \verb|8(r11)| in any instruction.  Using the
third item as the address of the data is just as easy with
\verb|@8(r11)| as the operand.  It is easy to see why one of the first
computers Forth was implemented on was the DEC PDP-11, which had many
of these same addressing modes.

\subsection{Methods of Threading}

The ``threaded'' in Threaded Interpretive Language comes from the
method used to implement the language, threading.  Threaded code
organizes the control of program flow into subroutine invocations
using a return stack to save the address to resume execution at after
each subroutine invocation.  There are several varieties of threaded
code, but only two will be considered here.

\subsubsection{Pointer Threading}

The most common method of implementing a TIL is to use {\em
  pointer}-threading.  With pointer-threading there are two different
kinds of words: secondaries and primitives.  A secondary word consists
of a list of points to other words, while a primitive consists of
machine code.  Secondaries are always called by other secondaries (the
outer loop that interprets the user's commands is a secondary), but
the secondaries can also call primitives.  Secondaries and primitives
are typically executed in the following way.  A special interpreter,
using called the {\em inner interpreter}, oversees execution  of all
words.  It maintains an {\em instruction register} which contains the
address of the next pointer in the current secondary word.  An inner
interpreter routine called {\bf next} extracts the address of the next
instruction, pointed to by the instruction register, and places it in
an {\em address register}, incrementing the instruction pointer to the
next pointer.  The routine {\bf run} then extracts the address pointed
to by the address register (which is whatever is in the first pointer
of the word pointed to by the pointer in the original word),
increments the address counter, and does a jump to the extracted
address.  If the word pointed to was a primitive, the address is that
of the first instruction in the machine code for the word, which is
then executed.  The last instruction in a primitive is a jump to the
{\em next} routine, which returns to the calling word.  If the word
pointed to was a secondary, the address is that of an inner
interpreter routine called {\bf colon}, which pushes the instruction
register contents on the return stack and moves the address register
contents  (the address of the second pointer in this new secondary) to
the instruction register and jumps to {\bf next}.  This effectively
nests down a level into the new secondary word.  The last pointer in a
secondary word is to the inner interpreter word {\bf semi}, which
simply pops the top address from the return stack and
%%% ??? Huh: should be "loads the address to the instruction register?
loads the address of the instruction register, thereby de-nesting one
level to the original secondary word.  See \cite[pages 18--20]{Start}
for more information on the inner interpreter.  (Note that on machines
with few registers the ``registers'' above would actually be memory
locations.)

This complicated pointer-threading scheme was developed for Forth in
the days of 8- and 16-bit processors, when typically only one register
could make use of stack instructions, the system stack register.
Since TILs make such heavy use of the data stack, for efficiency
reasons the system stack was used for the data stack, and another
method of linking subroutines had to be developed, that explained
above, which really simulates the normal system stack of a processor.
This implementation is very flexible and also produces very dense
code, but is unfortunately much slower than if the system stack had
been used for routine linkage because of the overhead of the innter
interpreter.\cite{CL}

\subsubsection{Subroutine Threading}

Since many modern processors allow any of several regeisters to be
used with stack instructions, another method can be used when
implementing a TIL, that of {\em subroutine}-threading, which is the
method I used in VTIL.  The VAX architecture allows any general
purpose register to control a stack, so the normal system stack can be
used for routine linkage.  With subroutine-threading a word consists
of a list of jump-to-subroutine or branch-to-subroutine instructions
instead of a list of pointers to words.  In pure subroutine-threading
a word consists {\em only} of the jump-to-subroutine instructions;
hwoever, it is often more useful (and just as easy) to allow other
machine language instructions to be intermixed with the
jump-to-subroutine instructions, which is what I have done in VTIL.
This allows more efficient code and has several other advantages that
will be discusses later in the section on improvements to VTIL.
Subroutine-threading is much faster than pointer-threading because
none of the overhead of the inner interpreter is involved.  It is also
considerably simpler to implement.  See \cite{CL} for information on a
commercial implementation of forth using this method.

\subsection{Text Interpreter}

Commands or source text, whether typed by the user at the terminal or
loaded from a text file, are parsed and acted on byh al oop, which is
called the text interpreter, or simply the interpreter.  The
interpreter is actually several conncted routines that are responsible
for either executing words, compiling words into the dictionary, or
pushing user entered data onto the data stack.  Wheter a word should
be compiled is determined by the value of the global variable
\verb|mode|, which is true (nonzero) if VTIL is in compile mode, and
false if it is execute mode.  The interpreter is only executed when
commands from the user are being read at the \verb|VTIL>| prompt; at
other times words are being directly executed, since they considt of
machine code.

Regardless of whether the word is to be executed or compiled the
interpret begins in the same way, by extracting the next blank
delimited token from the input stream and placing it at the end of the
dictionary (with actually placing it in the dictionary, which would
involve incrementing the dictionary pointer past the token) as a byute
contain the lenght of the token followed by the characters of the
token.  Next, the address of the {\em context} vocabulary, the
vocabulary in which words are searched for, is placed on the stack and
\word{?search} is called.  \word{?search} searches through the context
vocabulary for  word with a name maching the token.  If it is found
\word{?search} returns a true on the top of the stack and the address
of the header for the word under that.  If it is not found a false is
returned on the top of the stack.  If VTIL is in complie mode and the
word is not in the context vocabulary, \word{?search} also searches
the {\em compiler} vocabulary, which contains the words that can only
be in compile mode.  If it is found a true is returned on the top of
the stack and the address of the header for that word is returned
under that, just as normal, and the global variable \verb|state| is
set true, so that later on another routine will know that it is a
compile time word and execute it.  If it is not found, a false is
returned on the top of the stack, just as normal.

If the word was found the headerless routine \word{?execute} is
called.  First, the address of the code for that word is calculated
from the address of the header.  Then, if VTIL is in execute mode, a
jump to subroutine to the address on the top of the stack is done
(\verb|jsb @(r11)+|), and when that returns control branches to the
beginning of hte text interpreter loop.  However, if VTIL is in
compile mode a somewhat more complicated process must be done.  If
\verb|state| is true, the word is a compile-time word and is executd jast as
if VTIL were in execute mode.  If \verb|state| is false, the word has
to be compiled into the dictionary, which is done by
\word{compile-jsb}.  The machine code for the \verb|jsb| instruction
is moved into the dictionary along with the code for absolute
addressing mode, and immediately following that the address from the
top of the stack is moved into the dictionary.  After this control
branches to the beginning of the interpreter loop.

If a match was not found at all, the word \word{?number} is called.
\word{?nubmer} tries to convert the token into a number, according to
the current system radix (which can be changed to anything between two
and thirty-six).  If it can do so a true is returned on the top of the
stack and the number is returned under that.  If the token could not
be turned into a number a false is returned as the top of the stack.
Finally, the top of the stack is checked.  If it is false, the unknown
token is written out in an error message.  Control branches back to
the top of the interpreter loop.

Checking for stack underflow or overflow is also done each time
thorugh the interpreter.  The process of getting a token and either
executing, compiling, or pushing on the loop stack is continued until
the user is finished and VTIL is exited.

\subsection{Dictionary}

The dictionary is an area of contiguous memory which contains teh
data, code, and deaders for each defined word.  As new words are
defined the dictionary grows towards the top of memory.  The current
version of VTIL has allocated 16 kilobytes for expansion of the
dictionary, but the size of this area may be increased as desiered by
increasing a symbol in the VAX-Macro source code.

Every entry in the VTIL dictionary has the following fields:
\begin{description}
  \item[link] A longword that contains teh address of the header of
    the previous word in the dictionary.  The link field of the first
    word in the dictionary (and any independent vocabularies, such as
    the compiler  vocabulary) contains a zero, a null pointer, to
    indicate the end of the list.
  \item[length] A byte containing the length of the word's name.   The
    maximum name length is 80 characters.  During compilation of
    colon definitions, the high bit ({\em smudge bit}) of the length
    byte is set, preventing this definition of the word from being
    used recursively and allowing earlier definitions of this word to
    be used in defining this word.
  \item[name] {\em length} bytes containing the characters of the name
    of the word.
  \item[code] The executable code for this word.
  \item[data] The optional data for this word.
\end{description}

The current version of VTIL has these fields in each word in the
order of the preceding list; however, such an order is almost
completely arbitrary from the user's point of view (though not to
the implementer!), and could be switched, as long as words to
convert from the address of one field to the address of another
field are written.

Searching the dictionary is very simple.  Since the header of each
word consists of the link field followed by the name field, one
can simply step through the linked list and compare the token one
is searching for with the name of each word.  The search is
stopped when a match is found or the link to the next word is a
null pointer (zero).  Searches are always (and only, since
vocabularies are singly linked lists) done from the most recent
definition to the oldest definition.

\subsection{Compilation}

Compiling into the dictionary is a simple process in VTIL.  Register
nine is the {\em dictionary pointer}, which always points to the end of
the dictionary.  Compiling is simply a matter of copying the
information into the area of memory pointed to by the dictionary
pointer and incrementing the dictionary pointer over the new
information.  The VAX's autoincrement and autodecrement addressing
modes are extremely useful here, since they allow one to move data
directly into the dictionary and not have to worry about incrementing
the dictionary pointer.  So, compiling words often consist largely of
move instructions.  The difficult matter is figuring out what to
compile into the dictionary.

The simplest word is one that does nothing when it is executed.  It is
compiled by setting up the header for the word and putting a
\verb|rsb| immediately after the header as the first and only
instruction in the word.  Compiling a word that consists only of calls
to other words is done by copying a \verb|jsb| and the addressing mode
specifier for absolute addressing mode to the dictionary, followed by
the address of the code for the word being called, for each word that
is to be called.  Most of the compiling that is done is of these siple
types: compiling a header, compiling \verb|jsb| instructions, and
compiling \verb|rsb| instructions.  Really, any single instruction is
easy to compile.  The more difficult parts of compiling, as well as
the more interesting, come in dealing with control structures,
especially loops.  Following is a discussion of the
\word{if ... else ... endif} control structure, which will illustrate
the general ideas behind words the compile control structures.  The
dictionary pointer is \verb|R9| and the loop stack pointer is
\verb|R10|.  Words for the beginning of control structures generally
save data that the ending word must know on the loop stack instead of
the data stack, to avoid disturbing what the user may have placed on
the data stack.

Here is the VAX-Macro code for \word{if}:
\begin{verbatim}
if:
        movb    #op_tstl,(r9)+       ;; tstl
        movb    #adm_ainc_r11,(r9)+  ;; (r11)+
        movb    #op_bneq,(r9)+       ;; bneq
        movb    #^x03,(r9)+          ;; to (pc+3)
        movb    #op_brw,(r9)+        ;; brw
        movl    r9,-(r10)    ; save address of displacement
        movw    #0,(r9)+     ; save space for displacement
        movl    r9,-(r10)    ; save address to calculate
                             ; displacement from end of if
        rsb
\end{verbatim}

The first five lines of code compile the test of the top of the stack
and set up the branches to the true part and the false part.  The end
result is code in the dictionary equivalent to the following
VAX-Macro.
\newpage
\begin{verbatim}
        tstl    (r11)+  ; top of stack is true if nonzero
        bneq    10$     ; branch INTO true part
        brw     100$    ; branch OVER true part
10$:
\end{verbatim}

I used the \verb|bneq| to branch into the true part instead of just
having a \verb|beql| branch around it, allowing \word{if} statements
with more ethan 127 bytes of code in each branch, since \verb|brw|
uses a signed VAX memory word and \verb|beql| uses a signed byte
offset.  Since the length of hte true part will not be known until an
\word{else} or \word{endif} is found, the sixth line saves the address
of displacement for the jump over the true part on the loop stack
and the seventh line initializes the displacement to zero, saving
space for it in the dictionary.  Finally, the eight line saves on the
loop stack the address that the displacement over the true part will
be calculated from.

Here is the VAX-Macro code for \word{else}:
\begin{verbatim}
else:
        movb    #op_brw,(r9)+  ; branch over false part
        movl    r9,r0          ; save address of displacement
        movw    #0,(r9)+       ; save space for displacement
        movl    49,r1          ; save address to calculate
                               ; displacement from
; make branch to false part (fill in displacement in if)
        subl3   (r10)+,r9,r2   ; calculate displacement
        movw    r2,@(r10)      ; save displacement
        addl2   #4,r10   ; pop displacement address off stack
        movl    r0,-(r10)      ; save displacement address
        movl    r1,-(r10)      ; save address to displace from
        rsb                    ; end of else
\end{verbatim}

The first four lines of this code compile the branch over the false
part into the dictionary and save the address of the displacement and
the and the address to displace to in registers, since the top of the
loop stack has information that must be used now to fill in the
displacements from the \word{if}.  Lines seven through nine calculate
the displacement over the true part and save it in the code compiled
by the \word{if}.  Finally, the last two lines before the \verb|rsb|
save the displacement address and the address to displace from ont he
loop stack, so they will be preserved until the \word{endif}.

\newpage
Here is the VAX-Macro code for \word{endif}:
\begin{verbatim}
endif:
        subl3   (r10)+,r9,r0 ; calculate displacement
        movw    r0,@(r10)    ; save displacement
        addl2   #4,r10       ; pop displacement address
                             ; off loop stack
        rsb                  ; end of endif
\end{verbatim}

The first line calculates the displacement over the preceding branch
(which could either be the \word{if} or the \word{else} branch).
The second line saves that displacement in the code that was compiled
by either \word{if} or \word{else}.

\subsection {Error Handling}

Error handling in VTIL is simple but relatively robust, allowing the
user to remain in VTIL after all but the most serious errors.  Simple
errors, such as a non-existing file being specified to \word{load},
are caught by the word in question and indicated to the user.  Words
for reading input from the terminal truncate lines too long for their
buffers without telling the user.  More complex or more severe errors,
such as access violations, are generally caught by a condition handler
which signals them using the \verb|$PUTMSG_S| system service and then
unwinds the stack and restarts VTIL.  \verb|Ctrl/c|s are trapped by
an AST routine that signals the VTIL error ``Interrupted by user''.
This signal is then caught by the condition handler and an unwind and
restart of VTIL is performed as normal.

\section{Improvements}

After having implemented VTIL it is easy to see many ways in which it
could be improved.  Most of these are additional capabilities that
would make it easier for teh user to interact with VTIL or that would
allow programs to interact more fully with the normal VAX
environment.  Undoubtedly more work with the language will reveal
other things that also need to be improved (for instance, the current
string routines are clumsy in concept and execution).

Perhaps the most important thing to add to VTIL is some sort of file
access.  Currently, the only form of mass storage VTIL has access to
is the block file; however, named files of indefinite length are
usually easier to work with, and since VTIL runs on top of an
operating system the gains of efficiency that would accrue in the case
where VTIL itself is the operating system  and does its own block
input and output are lost in any case.  VTIL would therefore profit
from a set of words for opening, reading, writing, and closing normal
files.  The difficulty of this will be in figuring out how to do this
dynamically, rather than merely allocating a fixed number of file
control blocks, so that as many files can be open at one time as are
needed by the users application.  Ideally, all of VAX RMS would be
available to the user; this goal was accomplished in STOIC, which
allows access to RMS and most of the VAX System Services as well.

Another important enhancement would be allowing \word{load}s to be
nested.  This would allow the programmer to create words in small text
files which could be loaded singly to test the words independently,
but could be all linked together by nested \word{load}s when the
application is complete.  Keeping source code in small linked files
instead of a single huge file eases testing and debugging, since only
one file needs to be changed.

VTIL does not currently check to see if its control structures are
nested correctly in a definition.  VTIL doesn't care in the least if
the programmer forgets an \word{endif} or \word{loop}, though the
programmer may care a great deal!  While this is not as serious a
problem as it might seem, since VTIL words are short, nesting is
rarely deep, and finding such problems is therefore usually simple, it
would be nicer if VTIL itself could find these problems.  One method
would be during compilation for each control structure that begins a
block of code to push a unique value onto the loop stack and each
control structure that ends a block of code to check that the top fo
the loop stack was indeed put there by the corresponding word.  If it
was not, a word would be called that would step down through the loop
stack listing the improperly closed control structures until there
were no more left and then abort compilation.

Another area where VTIL needs improvement is in its interactive
execution of words; the inability to use any of the loop or control
structures and the need for two different versions of some words (like
\word{ascii}, which pushes the ASCII value of the first character of
the next token to the stack, and \word{[ascii]}, which compiles the
ASCII code of the first character of the next token into the current
word) for both compile-time and run-time is annoying and confusing.
This could be eliminated by using a {\em compile buffer}, into which
words that are executed interactively are compiled instead of
immediately executed.  At the end of a line of commands, the
interpreter could place a return from subroutine instruction at the
end of the buffer and perform a jump to subroutine to the beginning of
the buffer.  Since the same process of compiling now goes on
regardless of whether one is executing or interpreting these words
there will be no need to have two versions of words.

This would also allow the use of control structures in interactive
mode.  Currently, all control structure words are found only when
inside a colon definition because they compile things into the
dictionary and disturb the stacks.  If the compile buffer were added,
though, they would be able to perform their normal compiling tasks
into the compile buffer, allowing them to be used interactively.  In
order to keep a loop from being executed before it is closed by the
word to end that particular loop, each opening loop word would
increment a {\em loop depth} count and each closing word would
decrement the counter.  The compile buffer would be executed only when
the loop counter is zero; that is, only at the end of a line when
there are no unclosed loops.  Otherwise, the interpreter will get
another line of commands from the user.  This method will also
eliminate the need for a separate compiler vocabulary.  Unfortunately,
this scheme would complicated the task of defining words, especially
colon definitions.

An enhancement that would allow more flexibility in the execution and
compilation of words, as well as incrase efficiency of the compiled
code is an {\em attribute byte} for each word.  This byte would
contain a value that would tell the interpreter how to deal with the
word; for instance, a word that needed to be executed immediately,
such as \word{do}, which starts a \word{do ... loop}, would have the
attribute {\em jump-to-me}, while another word like \word{+rot} which
should be compiled into the compile buffer or definition would have
the attribute {\em compile-jsb-to-me}.  Several other attributes for
different purposes are possible.

If a length for each word were also stored in its header, an attribute
{\em expand-me} could be created.  Words with the {\em expand-me}
attribute are copied into the compile buffer instead of having a jump
to subroutine constructed to them.  This inline expansion of code
would make VTIL much more efficient without increasing the code size
much, if any.  Consider the code for the word \word{+}:
\begin{verbatim}
i_plus:
        addl2   (r11)+,(r11)
        rsb
\end{verbatim}

If this code (excluding the rsb) were expanded into each definition
where it was used, it would faster and smaller in code size than
compiling a \verb|jsb| to it.

\section{Conclusions}

VTIL has proved to be a very interesting exercise in language
implementation.  The language itself is a good general purpose
programming language, with its only major limitations being in access
to files, a limitation that can be overcome.  VTIL, compared to a
normal interpreter or compiler, takes relatively little time to
implement, which is valuable in cases where resources are limited but
a more powerful programming environment is needed.  It is a very
flexible language that the user can adapt to their specific
application.  VTIL is easily extended to handle complex problems.
Debugging VTIL programs is very easy because programs are separated
into small pieces (words) that can be easily tested and modified
separately.  Testing a word is simple because of the interactive
nature of VTIL and the simplicity of the stack-based implicit
parameter passing that VTIL uses for communication between words.
Although it was implemented in VAX-Macro on a VAX, VTIL could just as
easily be implemented on most machines.  Once the kernel and compiler
vocabularies have been implemented on another machine VTIL programs
from a VAX could be easily ported to the other machine, yet the
machine specific resources of each machine could be utilized fully if
needed.   There are also drawbacks to VTIL, however.  It uses many
concepts that are unfamiliar to a programmer used to more conventional
languages.  The postfix notation necessitated by the stack is alien
and difficult for many people to adjust to.  The greatly reduced
reliance on variables in calculation is also greatly discouraging to
traditional programmers.  It can be criticized as an untyped language
is not built in.  However, type-checking can be added if desired.
VTIL is obviously not as efficient as a good compiler, but it is much
more efficient than the typical interpreter.

VTIL has proved itself to be a reasonably powerful language that can
be implemented with relatively few man-hours.  The line editor which I
wrote in VTIL shows that it can be used for reasonable applications.
The extensibility of the language makes it ideal for developing ones
own programming language environment.  It embodies the interesting
characteristics that make TILs so different from more conventional
languages.  I have enjoyed this project immensely, and plan to
continue working with VTIL and improving it.

\newpage
\addcontentsline{toc}{section}{References}
\begin{thebibliography}{12}
  \bibitem{Start} Brodie, Leo.  {\it Starting Forth}.  2nd ed. Englewood
    Cliffs: Prentice-Hall, 1987.
  \bibitem{Think} Brodie, Leo.  {\it Thinking Forth}.  Englewood
    Cliffs: Prentice-Hall, 1984.
  \bibitem{CL} Chavez, Lori. ``A Fast Forth for the 68000.''  {\it
      Dr. Dobb's Journal}.   12.10 (1987): 32--36.
  \bibitem{ERA}  Eldridge, Rodrick A.  {\it VAX-11 fig-Forth Version
      2.0 and Glossary}.  Ames: Central Iowa Forth Interest Group.
    DECUS Program Library Package V-SP-52.
  \bibitem{EMS}  Ewing, Martin S.  {The Caltech FORTH Manual.}
    Pasadena: California Institute of Technology, 1974 (Internal
    Report).
  \bibitem{HK}  Harris, Kim.  ``Forth Extensibility: or how to write a
    Compiler in Twenty-five Words or Less.''  {\it Byte}.  5.8 (1980):
    164--184.
  \bibitem{HR}  Hauck, Roger.  {\it Stoic (Stack Oriented Interactive
      Compiler)}.  Cambridge: Smithstonian Institute, 1980 (Internal
    Documentation).  DECUS Program Library Package V-SP-49.
  \bibitem{JJ}  James, John S.  ``What is Forth?  A Tutorial
    Introduction.''  {\it Byte}.  5.8 (1980): 100--126.
  \bibitem{LRG} Loeliger, R. G.  {\it Threaded Interpretive
      Languages}.  Peterborough: Byte Books, 1981.
  \bibitem{MCH} More, Charles H.  ``The Evolution of Forth, an Unusual
    Langauge.''  {\it Byte}.  5.8 (1980): 76-90.
  \bibitem{RTGW} Ritter, Terry, and Gregory Walker.  ``Varieties of
    Threaded Code for Language Implementation.''  {\it Byte}.  5.9
    (1980): 206--227.
\end{thebibliography}
\end{document}
